---
title: "Missing Mediterranean"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages, include=FALSE}
#install.packages("tidyr")
#install.packages("dplyr")
library(tidyr)
library(dplyr)
library(stringr)
library(data.table)
library(ggplot2)

#install.packages("table1")
library(table1)

#install.packages("knitr")
library(knitr)
```


## Cleaning IOM Missing Migrants Project data

For information on the Missing Migrants Project and their methodology, see [repository](https://github.com/stklem/med_migration/blob/master/README.md).

**Step 1:** load all Missing Migrants data and filter to include only the European and Mediterranean region

```{r Input missing med}
missing <- data.frame(fread("input/MissingMigrants-Global-2019-11-04T17-25-00.csv"))
  
missing_med <- missing %>% filter(Region.of.Incident == "Europe" |
                                    Region.of.Incident == "Mediterranean")
```

**Step 2:** Exploring and Addressing NAs
```{r Check NAs, eval=FALSE}
lapply(missing_med, function(x) length(which(is.na(x)==TRUE)))
lapply(missing_med, function(x) length(which(x == "")))
```

Observations: 

1. If the # dead or # missing were 0, the methods said they were listed as blank rather than 0. We will confirm that the sum of the dead and missing is equal to the total dead and missing column, then replace the NAs with 0.

```{r NAs for # dead and # missing}
# looking @ NAs for Number.Dead and Minimum.Estimated.Number.of.Missing
missing_med %>% 
    filter(is.na(Number.Dead) == TRUE) %>%
    summarise(mismatch = sum(which(Minimum.Estimated.Number.of.Missing != Total.Dead.and.Missing)))
  
missing_med %>% 
    filter(is.na(Minimum.Estimated.Number.of.Missing) == TRUE) %>%
    summarise(mismatch = sum(which(Number.Dead != Total.Dead.and.Missing)))

# for all incidents where the Number.Dead is NA, the minimum number of missing is equal to the value in and vice versa
# Now that that is confirmed, set NA for Dead and Missing to 0

missing_med <- missing_med %>% mutate_at(c("Number.Dead", "Minimum.Estimated.Number.of.Missing"),
                                        function(x) as.integer(str_replace_na(x, replacement = 0)))
```

2. If the number of survivors, females, males or children were unknown, they were left blank. We will leave these as NA. 

3. The only area of concern is the 147 NAs in Migration Route. Upon examination, this corresponds with the incidents that were in the "European" region. These must be manually added. The re-uploaded data includes:
+ Migration routes that were previously blank, based on location description and news articles
+ Added a column for "Sea or Land"

```{r manually add migration routes}
fwrite(missing_med, "output_data/missing_med_cleaned.csv")
missing_med <- data.frame(fread("output_data/missing_med_cleaned_manual.csv"))
```

**Step 3**: Condensing causes of death

The original cause of death variable has 72 answers. The brief variable has 12 levels
```{r}
missing_med <- missing_med %>% mutate(
  Cause.of.Death.Brief = as.factor(ifelse(str_detect(Cause.of.Death, "^Asphyxiation$|^Suffocation$|Asphyxiation, Drowning|Murdered, Asphyxiation|Presumed asphyxiation|Suffocation, Drowning|^Crushed$|Crushed, Asphyxiation"),"Suffocation",
                          ifelse(str_detect(Cause.of.Death,regex("Burn|Fire", ignore_case = TRUE)),"Burned",
                           ifelse(str_detect(Cause.of.Death, "Crushed, Drowning|^Drowning$|Drowning, Dehydration|Drowning, Suffocation|Presumed drowning|Presumed drowning, Hypothermia"),"Drowning",       
                            ifelse(str_detect(Cause.of.Death, "Cardiac arrest|Dehydration|Exhaustion|Exposure|Hyperthermia|Hypothermia"),"Exposure/Dehydration",
                             ifelse(str_detect(Cause.of.Death, "^Fall$|Fall into a mountain crevice|Head injury|Hit by car|Hit by train|Hit by truck|Hit by vehicle|Rockslide"), "Foot migration, accidental",       
                              ifelse(str_detect(Cause.of.Death, "Fall from border fence|Gassed"),"Indirect Violence",
                               ifelse(str_detect(Cause.of.Death, "Plane stowaway"),"Plane stowaway",
                                ifelse(str_detect(Cause.of.Death,"Hypoglycemia|Organ failure|Post-partum complications|Pulmonary edema|Sickness and lack of access to medicine"),"Sickness and lack of access to medicine",
                                 ifelse(str_detect(Cause.of.Death, "Suicide"), "Suicide",
                                  ifelse(str_detect(Cause.of.Death, "Crushed|Collision|Electrocution|Fall|Fell|Train|Vehicle"), "Transportation accident",
                                   ifelse(str_detect(Cause.of.Death, "Unknown"), "Unknown",
                                    ifelse(str_detect(Cause.of.Death, "Murder|Shot|Stabbed|Violence"), "Violence", NA))))))))))))))
```

**Step 4**: Converting to correct data types

```{r Convert data types}
# converting total dead and missing from string w/ comma to integer
missing_med$Total.Dead.and.Missing <- str_remove(missing_med$Total.Dead.and.Missing, ",")
missing_med$Total.Dead.and.Missing <- as.integer(missing_med$Total.Dead.and.Missing)

# convert region, cause of death, migration route, and geographical group to columns
factor_cols <- c("Region.of.Incident", "Cause.of.Death", "Migration.Route", "UNSD.Geographical.Grouping")
missing_med <- mutate_each(missing_med, "as.factor", factor_cols)

# parse dates
missing_med <- data.table(missing_med)
missing_med$Reported.Date <- as.Date(missing_med$Reported.Date, format = "%d-%b-%y")
missing_med$Year.Month <- format(missing_med$Reported.Date, "%Y-%m")
missing_med$Month.Year <- format(missing_med$Reported.Date, "%m/%y")
missing_med <- as.data.frame(missing_med)
```

**Step 5**: Summarizing by month and day

```{r summarize IOM}
summary_missing_day <- missing_med %>%
  group_by(Migration.Route, Reported.Date) %>%
  summarize(incidents = n(),
            dead = sum(Number.Dead, na.rm  = TRUE), 
            missing = sum(Minimum.Estimated.Number.of.Missing, na.rm  = TRUE), 
            total_dead_missing = sum(Total.Dead.and.Missing, na.rm  = TRUE))

summary_missing_month <- missing_med %>%
  group_by(Migration.Route, Year.Month) %>%
  summarize(incidents = n(),
            dead = sum(Number.Dead, na.rm  = TRUE), 
            missing = sum(Minimum.Estimated.Number.of.Missing, na.rm  = TRUE), 
            total_dead_missing = sum(Total.Dead.and.Missing, na.rm  = TRUE))
```

## Cleaning UNHCR Arrival Data
```{r clean UNHCR}
unhcr_arrival_country <- data.frame(fread("input/unhcr_arrivals_12_31.csv"))

# removing the disaggregation by land and sea
unhcr_arrival_country <- unhcr_arrival_country[, -which(str_detect(colnames(unhcr_arrival_country), "sea|land"))]

# tidying
unhcr_arrival_country$month <- formatC(unhcr_arrival_country$month, width=2, flag="0")
unhcr_arrival_country$Year.Month <- paste(unhcr_arrival_country$year, 
                                       unhcr_arrival_country$month, 
                                       sep = "-")
#removing unnecessary date variables
unhcr_arrival_country <- unhcr_arrival_country[, 4:9]
unhcr_arrival_country <- gather(data = unhcr_arrival_country, key = country, value = sum, -Year.Month)
unhcr_arrival_country$country <- str_remove(unhcr_arrival_country$country, "_all")

```

The UNHCR arrival info is by country, rather than migration routes like the IOM data. To compare, we must add a variable for the migration route. Spain is a part of the Western Mediterranean route, Greece and Cyprus are in the Eastern Mediterranean route, and  Italy and Malta are in the Central Mediterranean route. 

```{r UNHCR derive route}
unhcr_arrival_country <- unhcr_arrival_country %>% mutate(
  Migration.Route = if_else(str_detect(country, "italy|malta"), "Central Mediterranean", 
                  (if_else(str_detect(country, "greece|cyprus"), "Eastern Mediterranean",
                           "Western Mediterranean"))))

unhcr_arrival_route <- unhcr_arrival_country %>% 
  group_by(Migration.Route, Year.Month) %>%
  summarise(arrivals = sum(sum, na.rm  = TRUE))
```


## Joining IOM and UNHCR Data

```{r joining IOM and UNHCR}
summary_missing_month$Migration.Route <- as.character(summary_missing_month$Migration.Route)

summary_all_month <- right_join(summary_missing_month, unhcr_arrival_route, 
                                by = c("Migration.Route", "Year.Month"))

# When there were no incidents of death in a month, those values appeared as NA. Converting NAs to 0, to indicate that there were no deaths
summary_all_month <- data.frame(lapply(summary_all_month, function(x) replace_na(x, 0)))

# To facilitate easy visualization, we're adding individual variables for month and year
summary_all_month <- summary_all_month %>% mutate(
  Year = substr(as.character(Year.Month), start = 1, stop = 4),
  Month = substr(as.character(Year.Month), start = 6, stop = 7)
)

```

##Calculating Mortality
The key metric of our study is the mortality rate for migrants attempting to cross the Mediterranean. To calculate this, we must:
1. Add the total dead and missing to the total arrived, to determine the total number of people who attempted to cross
2. Divide total dead and missing by the total attempted and multiply by 1,000. This calculates Mortality rate per 1000 attempts

```{r calculating mortality}
summary_all_month <- summary_all_month %>% mutate(
  total_attempted = total_dead_missing + arrivals, 
  mortality = total_dead_missing / total_attempted * 1000)
```

To facilitate visualization, we also calculated the same values by year

```{r}
summary_all_year <- summary_all_month %>% 
  group_by(Migration.Route, Year) %>%
  summarise(
    incidents = sum(incidents),
    dead = sum(dead),
    missing = sum(missing),
    total_dead_missing = sum(total_dead_missing),
    arrivals = sum(arrivals))

summary_all_year <- summary_all_year %>% mutate(
  total_attempted = total_dead_missing + arrivals, 
  mortality = total_dead_missing / total_attempted * 1000)
```

```{r, echo=FALSE}
summary_all_month <- setDT(summary_all_month)
summary_all_year <- setDT(summary_all_year)
```

```{r, echo=FALSE}
fwrite(missing_med, "output_data/missing med final.csv")
fwrite(summary_all_month, "output_data/summary data by month.csv")
fwrite(summary_all_year, "output_data/summary data by year.csv")
```


